---
title: "Day 6: Guard Gallivant"
author: AndiF
date: 6 Dec 2024
output: 
  html_notebook: 
    code_folding: none
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(tictoc)
```

Read in the problem:

```{r}
dat <- read_lines("aoc06.txt")
mapmat <- str_split_fixed(dat, "", n = str_length(dat[1]))
```

Top left corner of the map:

```{r}
mapmat[1:10, 1:10]
```


### Part 1

Here are the moves:

```{r}
moves <- read.csv(text = "now,then,row_step,col_step
N,E,-1,0
S,W,1,0
W,N,0,-1
E,S,0,1")
moves
```

A dirty loop:

```{r}
work_map <- mapmat
cur_dir <- "N"
cur_pos <- which(work_map == "^", arr.ind = TRUE)
work_map[cur_pos] <- "X"

tic()
done <- FALSE
while (!done) {
  cur_move <- moves |>
    filter(now == cur_dir)
  
  new_pos <- matrix(c(NA, NA), nrow = 1, ncol = 2)
  new_pos[1] <- cur_pos[1] + cur_move$row_step
  new_pos[2] <- cur_pos[2] + cur_move$col_step
  
  if (!between(new_pos[1], 1, nrow(work_map)) ||
      !between(new_pos[2], 1, ncol(work_map))) {
    done <- TRUE  
  }
  else if (work_map[new_pos] %in% c(".", "X")) {
    work_map[new_pos] <- "X"
    cur_pos <- new_pos
  }
  else  {
    cur_dir <- cur_move$then
  }
}
toc()
```

Two bugs that wasted some time:

1. Forgot that the guard can walk over their own path again.

2. Needed to check that the new position was still on the map, rather than only current position.


```{r}
which(work_map == "X") |> length()
```

### Part 2

(Incomplete.)

Not sure about about this one... I'm going to try to brute force the example first.

```{r}
testmap_str <- c("....#.....",
".........#",
"..........",
"..#.......",
".......#..",
"..........",
".#..^.....",
"........#.",
"#.........",
"......#...")

testmap <- str_split_fixed(testmap_str, "", n = str_length(testmap_str[1]))
testmap
```



```{r}
has_loop <- function(the_map) {
  cur_dir <- "N"
  cur_pos <- which(the_map == "^", arr.ind = TRUE)
  blank_map <- matrix(0, nrow = nrow(the_map), ncol = ncol(the_map))
  been_here <- list(N = blank_map,
                    S = blank_map,
                    W = blank_map,
                    E = blank_map)
  
  done <- FALSE
  loopy <- FALSE
  while (!done) {
    been_here[[cur_dir]][cur_pos] <- been_here[[cur_dir]][cur_pos] + 1
    if (been_here[[cur_dir]][cur_pos] == 2) {
      loopy <- TRUE
      break
    }
    
    cur_move <- moves |>
      filter(now == cur_dir)
    
    new_pos <- matrix(c(NA, NA), nrow = 1, ncol = 2)
    new_pos[1] <- cur_pos[1] + cur_move$row_step
    new_pos[2] <- cur_pos[2] + cur_move$col_step
    
    if (!between(new_pos[1], 1, nrow(the_map)) ||
        !between(new_pos[2], 1, ncol(the_map))) {
      done <- TRUE
    }
    else if (the_map[new_pos] != "#") {
      cur_pos <- new_pos
    }
    else  {
      cur_dir <- cur_move$then
    }
  }
  
  loopy
}
```




```{r}
potentials <- which(testmap == ".", arr.ind = TRUE)
nrow(potentials)
```

```{r}
block_loops <- function(the_map, x, y) {
  the_map[x, y] <- "#"
  has_loop(the_map)
}
```


```{r}
tic()
res <- map2_lgl(potentials[,1],
                potentials[,2],
                \(x, y) block_loops(testmap, x, y))
toc()
```


```{r}
sum(res)
```

That worked, but it would be way too slow for the full problem (about a day), so the search needs to be much cleverer!

TBC


