---
title: "Day 3: Mull It Over"
author: AndiF
date: 3 Dec 2024
output: 
  html_notebook: 
    code_folding: none
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

Read in the problem:

```{r}
dat <- read_lines("aoc03.txt")
```

Line breaks aren't special, so paste the lot together:

```{r}
longmess <- paste(dat, collapse = "")
```

### Part 1

Regex time. Try with an easy test:

```{r}
easy_test <- "mul(2,4)ggsrjkgjdfmul(224,221)gielmul(1234,12)fskjd"
str_extract_all(easy_test, "mul\\([0-9]{1,3},[0-9]{1,3}\\)")
```

It worked. Now do it for _longmess_.

```{r}
mults <- str_extract_all(longmess, "mul\\([0-9]{1,3},[0-9]{1,3}\\)")[[1]]
mults |> head(20)
```

I'm going to let R parse each string.

```{r}
mul <- function(x,y) x*y

mult_one <- function(str) {
  str |> parse(text = _) |> eval()
}
```

Here's a test:

```{r}
mult_one("mul(6,7)")
```

The answer to part 1:

```{r}
map_int(mults, mult_one) |> sum()
```

### Part 2

The logic is, hike out anything between a "don't()" and a "do()", then run the code above on what remains. This also deals with consecutive do's automatically.

```{r}
get_do <- function(str) {
  doing_do <- TRUE
  dont_exp <- "don't\\(\\)"
  do_exp   <- "do\\(\\)"

  res  <- ""
  rest <- str

  while (str_length(rest) != 0) {
    if (doing_do) {
      loc <- str_locate(rest, dont_exp)
      
      if (!is.na(loc[1])) {
        res  <- paste0(res, str_sub(rest, 1, loc[1] - 1))
        rest <- str_sub(rest, loc[2] + 1)
      }
      else {
        res  <- paste0(res, str_sub(rest, 1))
        rest <- ""
      }
      
      doing_do <- FALSE
    }
    else {
      loc <- str_locate(rest, do_exp)
      
      if (!is.na(loc[1])) {
        rest <- str_sub(rest, loc[2] + 1)
      }
      else {
        rest <- ""
      }
      
      doing_do <- TRUE
    }
  }
  
  res
}
```

Give it a go:

```{r}
new_longmess <- longmess |> get_do()
```

A bit of copy and paste...

```{r}
mults <- str_extract_all(new_longmess, "mul\\([0-9]{1,3},[0-9]{1,3}\\)")[[1]]
map_int(mults, mult_one) |> sum()
```

And it works!
