---
title: "Day 2: Red-Nosed Reports"
author: AndiF
date: 2 Dec 2024
output: 
  html_notebook: 
    code_folding: none
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
```

The lines have different numbers of reports, so loading it in using a rawer method than the usual read.csv and friends:

```{r}
dat <- read_lines("aoc02.txt")
```


```{r}
head(dat)
```

That's a vector of strings. Next, split each string on the space and make a list of numeric vectors:

```{r}
repl <- strsplit(dat, " ") |> map(as.numeric)
head(repl)
```

### Part 1

The rules are:

* The levels are either all increasing or all decreasing.
* Any two adjacent levels differ by at least one and at most three.

Wrap in a function:

```{r}
test_report <- function(nums) {
  test_diffs <- diff(nums, lag = 1, difference = 1)
  all(test_diffs %in% 1:3) || all(-test_diffs %in% 1:3) 
}
```

The answer:

```{r}
res <- map_lgl(repl, test_report)
sum(res)
```


### Part 2

I'm going to brute force this one and try dropping elements even if it's unnecessary, as I don't have time to be clever 🙃

```{r}
robust_test_report <- function(nums) {
  if (test_report(nums)) {
    TRUE
  }
  else {
    res <- FALSE
    for (i in 1:length(nums)) {
      res <- test_report(nums[-i])
      if (res)
        break
    }
    res
  }
}
```

And it worked:

```{r}
robust_res <- map_lgl(repl, robust_test_report)
sum(robust_res)
```




