---
title: "Day 5: Print Queue"
author: AndiF
date: 5 Dec 2024
output: 
  html_notebook: 
    code_folding: none
---

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(stringr)
library(tictoc)
```

Read in the problem:

```{r}
dat <- read_lines("aoc05.txt")
```

Split into the rules and pages:

```{r}
split_i <- which(dat == "")
rules_raw <- dat[1:(split_i - 1)]
pages_raw <- dat[(split_i + 1):length(dat)]
```

Take a look:

```{r}
head(rules_raw)
```

```{r}
head(pages_raw)
```

Make a nice dataframe of rules:

```{r}
rules <- rules_raw |>
  str_split_fixed("\\|", 2) |>
  as.data.frame() |>
  rename(first = V1, second = V2)
head(rules)
```

Make a nice list of pages:

```{r}
pages <- strsplit(pages_raw, ",")
head(pages)
```


### Part 1

Test one rule on one vector of pages:

```{r}
test_rule <- function(first, second, pages) {
  first_i  <- which(pages == first)
  second_i <- which(pages == second)
  
  !is.unsorted(c(first_i, second_i))
}
```

Test all the rules on one vector of pages:

```{r}
test_all_rules <- function(page_vec) {
  map2_lgl(rules$first, rules$second,
           \(f, s) test_rule(f, s, page_vec)) |>
    all()
}
```

Test all the rules on all the pages:

```{r}
test_all_pages <- function(pages_list) {
  map_lgl(pages_list, test_all_rules)
}
```


```{r}
tic()
res <- test_all_pages(pages)
toc()
```

Look up the print runs that satisfy the rules:

```{r}
sorted_pages <- pages[res]
```

Sum the middle pages:

```{r}
mid_val <- function(vec) {
  middle <- (1 + length(vec)) / 2
  vec[middle] |> as.numeric()
}

map_int(sorted_pages, mid_val) |> sum()
```

### Part 2

Get the unordered pages:

```{r}
broken <- pages[!res]
head(broken)
```


The following section is brought to you by thinking, "Surely there's a graph algorithm for this", and also some wishful thinking about the completeness of the rules (prompted by the existence of a unique middle page).


```{r message=FALSE, warning=FALSE}
library(igraph)
```


```{r}
sort_pages <- function(vec) {
  rules |>
    filter(first %in% vec & second %in% vec) |>
    as.matrix() |>
    graph_from_edgelist() |>
    topo_sort() |>
    names()
}
```


```{r}
tic()
broken |>
  map(sort_pages) |>
  map_int(mid_val) |>
  sum()
toc()
```




